On shift equivalence for non-linear rules\thanks{Research partially supported by the MIUR PRIN 2017FTXR7S ``IT-MaTTerS''.}

### 2023-07-15 GraphShift/GraphShift.tex
added directory

\title{Ideas}

@@

\subsection*{Shift equivalence for non-right linear}

Sequential independence is a key ingredient in the concurrency theory of DPO graph transformation, since it states when consecutive rewriting steps can be switched, thus 
identifying them as causally unrelated. The associated theory of shift equivalence on rewriting sequences is well-understood for linear rules, including its 
connection with standard models for concurrent systems such as event structures. 
%
This is not so for rules that are only left-linear, i.e., such that
their right-hand side may merge graph items.

In particular, if one just take the definition with general
interchange pairs, undesidered things might happen. In particular one
can have shift-equivalent derivations where the permutation is the
identity but which are non isomorphic (e.g., in GraphShift, page 1-2
$\rho_1 \shifteq \rho_3$, but $\rho_1 \equiv^{a} \rho_3$.

We want a notion of shift-equivalence which ensures some properties
which are indispensable to allow to interpret derivations up-to shift
as equivalent concurrent computations which differ only for the fact
that independent steps have been moved around. When moving around steps, we should not loose track of the identity of the steps: this is essential to be able to have an event structure semantics with a clearly defined notion of event.

Hence we assume to deal
with a notion of shift-equivalence between derivations of the kind
$\rho \shifteq_\sigma \rho'$ where $\rho, \rho'$ consist of $n$
rewriting steps, $\sigma: \interval{n} \to \interval{n}$ is a
permutation (the idea being that $\rho$ is obtained from $\rho'$
pemutating the steps according to $\sigma$) with the following
properties:

\begin{itemize}
\item \textbf{Property 0}. Shift equivalence is an equivalence, in particular it is symmetric. I.e., if  $\rho \shifteq_\sigma \rho'$ then $\rho' \shifteq_{\sigma^{-1}} \rho$;

\item \textbf{Property 1}. Derivations which are shift-equivalent via the
  identity permutation are isomorphic: if $\rho \shifteq \rho'$ then
  $\rho \equiv^a \rho'$;

\item \textbf{Property 2}. Shift equivalence between adjacent rewriting steps
  is a property of the steps, invariant for permutation according to
  shift, i.e., if two steps are independent in some derivation $\rho$
  and we can prove that $\rho$ is shift-equivalent to some $\rho'$
  where the same steps are again consecutive, but possibly in a
  different position, then the steps are again independent, and also
  the converse holds. Formally, if $\rho \shifteq_\sigma \rho'$ with
  $\rho, \rho'$ consist of $n$ rewriting steps and for some
  $i \in \interval{n-1}$, $\sigma(i)+1 = \sigma(i+1)$, the the steps
  $i, i+1$ in $\rho$ are shift-independent iff $\sigma(i)$ and
  $\sigma(i)+1$ are so in $\rho'$.
\end{itemize}

\textbf{Question.} Taking the standard shift, with generic interchange pair, a a counterexample to property 1, is the one mentioned above.  Do we have a counterexample to property 2? Not clear to me.

The above discussion should lead to

\begin{enumerate}
\item The conditions proposed by Andrea for a good notion of shift (in
  the draft, GraphShift, Def. 3, cond (1)-(5), with the addition of
  Lemma 12).

  \textbf{Questions}: Can we reduce the set of axioms? Do we need something else relating abstraction equivalence (iso of derivation) and shift? Relation with symmetries?

\item For left-linear rules, the proposed notion of shift equivalence
  requires: uniqueness of the interchange pair and symmetry. Show that
  this satisfies the above

\item Apparently, the satisfaction of the properties above has not
  been explicitly observed for the case of linear rules. If so, it is
  worth to remark it.

\item Does the theory above apply to other cases. E.g., if one wants a
  proper notion of causality (avoiding or-causality) a possibility
  seems to be to restrict the applicability condition for rules, to
  prevent fusions to act on already merged items (see section V.D in
  the arxiv). By the way, note that the satisfaction of the conditions
  on shift depends not only on the notion of shift, but also on the
  notion of rewriting. In this case we expect to have a prime event
  structure (bit this has to do with the second part of the work).

\end{enumerate}

\subsection*{Event structures for non-right linear}

Ideally, one would like to be able to work with an abstract notion of
shift-equivalence and just using its properties as singled out in the
previous part derive a weak prime domain or connected event structure
for it. This is probably difficult, but one might need to add some
futher properties to be proved for the specific notion of shift at
hand.

For this part, relevant material is section V of Arxiv. We use abstraction equivalence (iso of derivations) $\equiv^{a}$ (def. 36), shift-equivalence $\shifteq_\sigma$ (Def. 38, traking the permutation) and their join $\equiv^c_\sigma$.

For linear rules on graphs it is known that two derivations are
related by $\rho \equiv^c_\sigma \rho'$ iff $\sigma$ is a consistent
permutation (def. 40, arxiv). Rough;y the idea is to take the colimit
and consider the rewriting steps over the colimit to deduce
dependencies. This was used in many places in the proofs, but for
left-linear rules the proof does not go through (only the implication
``if $\rho \equiv^c_\sigma \rho'$ then $\sigma$ is a consistent
permutation'' (Lemma 17(1)) seems to hold as the original proof do not
to rely on linearity, this should be double-checked, but the converse
is false). In general, the hope is to get rid of using consistent
permutation and use instead more asbtract properties of
shift-equivalence as discussed in the first part.

The general idea is to show that considering derivations are arrows from the source to the target graph, then the slice from the initial graph $\slice{[G_s]}{\tr{\mathcal{G}}}$ should be a pre-order, whose ideal completion is a weak prime domain.

Relevant technical lemma in Arxiv are:

\begin{itemize}

\item
  \textbf{Lemma 17}. Points (2) and (3) should be hopefully consequences of the general properties of shift in the previous part and they play a role in proving that $\slice{[G_s]}{\tr{\mathcal{G}}}$ is a pre-order (Prop. 14)

\item
  \textbf{Lemma 19}.
  (1) says that if $\psi; \psi_1 \equiv^c_\sigma \psi'; \psi_1'$ and the permutation $\sigma$ sends $n$ steps of $\psi_1$ to $\psi'$ then such steps are not in $\psi$ and they will need to be in every $\phi_2$ such that $\psi; \phi_2 \equiv^c_\sigma \psi'; \phi_2'$. In particular there is $\phi_2$ consisting of exactly $n$ steps which allows for commutation. This allows essentially to deduce existence of minimal upper bounds, al with the same length. 
  Then (2) shows that all minimal upper bounds are isomorphic, so that we have a sup.

  Again, the hope is to obtain this from the abstract properties of
  shift.

\item \textbf{Proposition 15} shows that $\ideal{\poset{\mathcal{G}}}$
  is a domain, i.e., algebraic finitary (see Def. 6), while
  \textbf{Theorem 6} prove that it is weak prime.

  Here I think that the proofs are draft modifications of the original ones, changed after discovering the property on consistent permutation to be wrong. But they are drafty, to be redone maybe keeping some ideas.

\subsection*{Event structures for non-right linear, but jointly mono matches}

This is the idea in Section V.D. As mentioned above, ideally, if the proof in the previous part were based just on abstract properties of shift, one could hope of proving that shift in this context has the right properties, deduce that the construction produces a weak prime domain.

Then, one could observe that in this specific context actually all weak primes are primes an thus deduce that this provides a prime algebraic domain/pes semantics of fusion.

\keywords{Shift equivalence, non-linear rules, DPO rewriting.}

\section{Introduction}
\label{se:graphs}

Shortly, shift equivalence can be defined as the smallest equivalence relation on derivations, i.e. sequences of rewriting steps,
that can be obtained by the iterated switch of sequentially 
independent steps. A crucial observation, which represented our starting point, is that the notion of sequential independence traditionally
adopted for linear graph transfomation systems is not adequate when we
allow rules to be non-right linear, i.e., in the presence of
fusions. Consider the rules in Fig.~\ref{fi:rules}, typed over the type graph $T$,
where numbers are used to represent the mappings from the interface to the left- and
right-hand sides.

@@

Now, let us consider an initial graph with a single node. We can then obtain
the derivation $\rho_1 = \delta_1; \delta_2; \delta_3$ in Fig.~\ref{fi:der1}, where $\delta_2$ and $\delta_3$ are
sequential independent according to the standard notion.

@@

The fact that if we switch two steps and then switch them back, we obtain a derivation that is not equivalent to the original one has unpleasant consequences. For instance, if we restrict $\rho_1'$ and $\rho_3'$ to the first two steps, we get non-equivalent derivations $\delta_1;\delta_2 \not\shifteq \delta_1;\delta_2'$, but $\rho_1 \shifteq[\sigma] \rho_3$, with $\sigma$ mapping $\delta_1$ to $\delta_1$ and $\delta_2$ to $\delta_2'$.
%
The lack of such closure with respect to composition does not allow to obtain a causal semantics for graph transformation systems.
More precisely, it does  not allow to associate an event structure to a derivation.
%
Thus, the notion of sequential independence has to be refined. The point is that in the associated shift equivalence we want to forget who made the fusion, 
but we should not forget (as it happens in the example) that some fusions occurred.

@@

Shift equivalence is made stricter by asking that two steps are
sequentially independent if the result of their switch is uniquely
determined. Intuitively, when there are several way of performing the
exchange, it means that the first step has executed some fusions that
the second step is using in an essential way, hence they should not
considered independent. Formally, this boils down to require that 
in the definition of sequential independence the interchange pair is unique.

The paper builds on this intuition: it introduces this stricter notion of sequential independence
(Section~xxx) and prove that the associated theory of shift equivalence on derivations is 
well-behaved with respect to the concatenation of rewriting steps (Section~xxx).
Finally, it outlines how the relationship with event structure works, wrapping up the paper
with comparison with related woks and hinting at further direction of research (Section~xxx).

\section{Preliminaries [to be redone]}

@@

\begin{proof}
  To be done. The fact that sequential independence coincides with
  independence for contiguous derivations is easy. The fact that it coincides when derivations are not contiguous is false (the approach based on defining independence for non-contiguous derivation is not working, but the results should be true ...)
\end{proof}

@@

  Moreover, let $\upsilon_1, \upsilon_2$ be adjacent
  transpositions and $\sigma' = \upsilon_1; \ldots, \upsilon_m$ a
  minimal length decomposition of $\sigma'$ in terms of adjacent
  transpositions (that thus operate on inversions). Then, for all
  $h \in \interval{m}$ with $\upsilon_h = \transp{l}[l+1]$ there exists
  $k \in \interval{m}$ such that
  $\upsilon_k = \transp{\upsilon_{1,h-1}^{-1};v_1;
    \upsilon_{1,k-1}(l)}[\upsilon_{1,h-1}^{-1};v_1; \upsilon_{1,k-1}(l+1)]$.
\end{lemma}

>>>>

\title{On graphs and shift}

@@

\section{Domain and event structure semantics for graph rewriting}
\label{se:graphs}

@@

A crucial observation is that the notion of shift equivalence normally
adopted for linear graph rewriting systems is not adequate when we
allow rules to be non-right linear, i.e., in the presence of
fusions. Consider the rewriting system in Fig.~\ref{fi:rules},
where numbers are used to represent the mappings from the interface to the left- and
right-hand sides.
%
Then, starting from an initial graph with a single node, we can obtain
the derivation $\rho = \delta_1; \delta_2; \delta_3$ in Fig.~\ref{fi:der1}, where $\delta_2$ and $\delta_3$ are
sequential independent according to the standard notion.

@@

The fact that if we shift two steps and then switch them back, we can obtain a derivation that is not equivalent to the original one has unpleasant consequences. For instance, if we restrict $\rho_1'$ and $\rho_3'$ to the first two steps, we get non-equivalent derivations $\delta_1;\delta_2 \not\shifteq \delta_1;\delta_2'$, but $\rho_1 \shifteq[\sigma] \rho_3$, with $\sigma$ mapping $\delta_1$ to $\delta_1$ and $\delta_2$ to $\delta_2'$.

The notion of shift equivalence has to be refined: the point is that we want to forget who made the fusion, but we should not forget (as it happens in the example) that some fusions occurred.

@@

Shift equivalence is made stricter by asking that two steps are
shiftable only when the result of the shift is uniquely
determined. Intuitively, when there are several way of performing the
exchange, it means that the first step has performed some fusions that
the second step is using in an essential way, hence they should not
considered independent. Formally, in the definition of sequential
independence (Fig.~\ref{fi:strongseq}) we ask that the independence pair $\langle i_1, i_2\rangle$ is unique.

@@

\begin{proof}
  To be done. The fact that sequential independence coincides with
  independence for contiguous derivations is easy.
\end{proof}

@@

  Moreover, let $\upsilon_1, \upsilon_2$ be adjacent
  transpositions and $\sigma' = \upsilon_1; \ldots, \upsilon_m$ a
  minimal length decomposition of $\sigma'$ in terms of adjacent
  transpositions (that thus operate on inversions). Then, for all
  $h \in \interval{m}$ with $\upsilon_h = \transp{l}[m]$ there exists
  $k \in \interval{m}$ such that
  $\upsilon_k = \transp{\upsilon_{1,h-1}^{-1};v_1;
    \upsilon_{1,k-1}(l)}[\upsilon_{1,h-1}^{-1};v_1; \upsilon_{1,k-1}(m)]$.
\end{lemma}

### 2023-07-15 GraphShift/icgt2023.tex
added directory

\title{Ideas}

@@

\subsection*{Shift equivalence for non-right linear}

Sequential independence is a key ingredient in the concurrency theory of DPO graph transformation, since it states when consecutive rewriting steps can be switched, thus 
identifying them as causally unrelated. The associated theory of shift equivalence on rewriting sequences is well-understood for linear rules, including its 
connection with standard models for concurrent systems such as event structures. 
%
This is not so for rules that are only left-linear, i.e., such that
their right-hand side may merge graph items.

In particular, if one just take the definition with general
interchange pairs, undesidered things might happen. In particular one
can have shift-equivalent derivations where the permutation is the
identity but which are non isomorphic (e.g., in GraphShift, page 1-2
$\rho_1 \shifteq \rho_3$, but $\rho_1 \equiv^{a} \rho_3$.

We want a notion of shift-equivalence which ensures some properties
which are indispensable to allow to interpret derivations up-to shift
as equivalent concurrent computations which differ only for the fact
that independent steps have been moved around. When moving around steps, we should not loose track of the identity of the steps: this is essential to be able to have an event structure semantics with a clearly defined notion of event.

Hence we assume to deal
with a notion of shift-equivalence between derivations of the kind
$\rho \shifteq_\sigma \rho'$ where $\rho, \rho'$ consist of $n$
rewriting steps, $\sigma: \interval{n} \to \interval{n}$ is a
permutation (the idea being that $\rho$ is obtained from $\rho'$
pemutating the steps according to $\sigma$) with the following
properties:

\begin{itemize}
\item \textbf{Property 0}. Shift equivalence is an equivalence, in particular it is symmetric. I.e., if  $\rho \shifteq_\sigma \rho'$ then $\rho' \shifteq_{\sigma^{-1}} \rho$;

\item \textbf{Property 1}. Derivations which are shift-equivalent via the
  identity permutation are isomorphic: if $\rho \shifteq \rho'$ then
  $\rho \equiv^a \rho'$;

\item \textbf{Property 2}. Shift equivalence between adjacent rewriting steps
  is a property of the steps, invariant for permutation according to
  shift, i.e., if two steps are independent in some derivation $\rho$
  and we can prove that $\rho$ is shift-equivalent to some $\rho'$
  where the same steps are again consecutive, but possibly in a
  different position, then the steps are again independent, and also
  the converse holds. Formally, if $\rho \shifteq_\sigma \rho'$ with
  $\rho, \rho'$ consist of $n$ rewriting steps and for some
  $i \in \interval{n-1}$, $\sigma(i)+1 = \sigma(i+1)$, the the steps
  $i, i+1$ in $\rho$ are shift-independent iff $\sigma(i)$ and
  $\sigma(i)+1$ are so in $\rho'$.
\end{itemize}

\textbf{Question.} Taking the standard shift, with generic interchange pair, a a counterexample to property 1, is the one mentioned above.  Do we have a counterexample to property 2? Not clear to me.

The above discussion should lead to

\begin{enumerate}
\item The conditions proposed by Andrea for a good notion of shift (in
  the draft, GraphShift, Def. 3, cond (1)-(5), with the addition of
  Lemma 12).

  \textbf{Questions}: Can we reduce the set of axioms? Do we need something else relating abstraction equivalence (iso of derivation) and shift? Relation with symmetries?

\item For left-linear rules, the proposed notion of shift equivalence
  requires: uniqueness of the interchange pair and symmetry. Show that
  this satisfies the above

\item Apparently, the satisfaction of the properties above has not
  been explicitly observed for the case of linear rules. If so, it is
  worth to remark it.

\item Does the theory above apply to other cases. E.g., if one wants a
  proper notion of causality (avoiding or-causality) a possibility
  seems to be to restrict the applicability condition for rules, to
  prevent fusions to act on already merged items (see section V.D in
  the arxiv). By the way, note that the satisfaction of the conditions
  on shift depends not only on the notion of shift, but also on the
  notion of rewriting. In this case we expect to have a prime event
  structure (bit this has to do with the second part of the work).

\end{enumerate}

\subsection*{Event structures for non-right linear}

Ideally, one would like to be able to work with an abstract notion of
shift-equivalence and just using its properties as singled out in the
previous part derive a weak prime domain or connected event structure
for it. This is probably difficult, but one might need to add some
futher properties to be proved for the specific notion of shift at
hand.

For this part, relevant material is section V of Arxiv. We use abstraction equivalence (iso of derivations) $\equiv^{a}$ (def. 36), shift-equivalence $\shifteq_\sigma$ (Def. 38, traking the permutation) and their join $\equiv^c_\sigma$.

For linear rules on graphs it is known that two derivations are
related by $\rho \equiv^c_\sigma \rho'$ iff $\sigma$ is a consistent
permutation (def. 40, arxiv). Rough;y the idea is to take the colimit
and consider the rewriting steps over the colimit to deduce
dependencies. This was used in many places in the proofs, but for
left-linear rules the proof does not go through (only the implication
``if $\rho \equiv^c_\sigma \rho'$ then $\sigma$ is a consistent
permutation'' (Lemma 17(1)) seems to hold as the original proof do not
to rely on linearity, this should be double-checked, but the converse
is false). In general, the hope is to get rid of using consistent
permutation and use instead more asbtract properties of
shift-equivalence as discussed in the first part.

The general idea is to show that considering derivations are arrows from the source to the target graph, then the slice from the initial graph $\slice{[G_s]}{\tr{\mathcal{G}}}$ should be a pre-order, whose ideal completion is a weak prime domain.

Relevant technical lemma in Arxiv are:

\begin{itemize}

\item
  \textbf{Lemma 17}. Points (2) and (3) should be hopefully consequences of the general properties of shift in the previous part and they play a role in proving that $\slice{[G_s]}{\tr{\mathcal{G}}}$ is a pre-order (Prop. 14)

\item
  \textbf{Lemma 19}.
  (1) says that if $\psi; \psi_1 \equiv^c_\sigma \psi'; \psi_1'$ and the permutation $\sigma$ sends $n$ steps of $\psi_1$ to $\psi'$ then such steps are not in $\psi$ and they will need to be in every $\phi_2$ such that $\psi; \phi_2 \equiv^c_\sigma \psi'; \phi_2'$. In particular there is $\phi_2$ consisting of exactly $n$ steps which allows for commutation. This allows essentially to deduce existence of minimal upper bounds, al with the same length. 
  Then (2) shows that all minimal upper bounds are isomorphic, so that we have a sup.

  Again, the hope is to obtain this from the abstract properties of
  shift.

\item \textbf{Proposition 15} shows that $\ideal{\poset{\mathcal{G}}}$
  is a domain, i.e., algebraic finitary (see Def. 6), while
  \textbf{Theorem 6} prove that it is weak prime.

  Here I think that the proofs are draft modifications of the original ones, changed after discovering the property on consistent permutation to be wrong. But they are drafty, to be redone maybe keeping some ideas.

\subsection*{Event structures for non-right linear, but jointly mono matches}

This is the idea in Section V.D. As mentioned above, ideally, if the proof in the previous part were based just on abstract properties of shift, one could hope of proving that shift in this context has the right properties, deduce that the construction produces a weak prime domain.

Then, one could observe that in this specific context actually all weak primes are primes an thus deduce that this provides a prime algebraic domain/pes semantics of fusion.

@@

The fact that if we switch two steps and then switch them back, we obtain a derivation that is not equivalent to the original one has unpleasant consequences. For instance, if we restrict $\rho_1'$ and $\rho_3'$ to the first two steps, we get non-equivalent derivations $\delta_1;\delta_2 \not\shifteq \delta_1;\delta_2'$, but $\rho_1 \shifteq[\sigma] \rho_3$, with $\sigma$ mapping $\delta_1$ to $\delta_1$ and $\delta_2$ to $\delta_2'$.
%
The lack of such closure with respect to composition does not allow to obtain a causal semantics for graph transformation systems.
More precisely, it does  not allow to associate an event structure to a derivation.
%
Thus, the notion of sequential independence has to be refined. The point is that in the associated shift equivalence we want to forget who made the fusion, 
but we should not forget (as it happens in the example) that some fusions occurred.

@@

>>>>

\title{On shift equivalence for non-linear rules\thanks{Research partially supported by the MIUR PRIN 2017FTXR7S ``IT-MaTTerS''.}
}

\author{Paolo Baldan\inst{1}, Andrea Corradini\inst{2}, Fabio Gadducci\inst{2}
} 
	\institute{Department of Mathematics, University of Padua, Italy \\
		\email{baldan@math.unipd.it}
		\and Department of Computer Science, University of Pisa, Italy \\
		\email{\{andrea.corradini, fabio.gadducci\}@unipi.it}
		}

\titlerunning{On shift equivalence for non-linear rules}
\authorrunning{Baldan, Corradini, and Gadducci, and Santini}

@@

\begin{abstract}
Sequential independence is a key ingredient in the concurrency theory of DPO graph transformation, since it states when consecutive rewriting steps can be switched, thus 
identifying them as causally unrelated. The associated theory of shift equivalence on rewriting sequences is well-understood for linear rules, including its 
connection with standard models for concurrent systems such as event structures. 
%
This is not so for rules that are only left-linear, i.e., such that their right-hand side may merge graph items. More precisely, the correspondence with
event structures fails, since merging may hinder the causal dependency between rewriting steps.
%
We argue that a stricter notion of sequential independence is needed for such rules, and we prove that our proposal is well-behaved with respect to the concatenation of
rewriting steps.
\end{abstract}

%We extend Soft Concurrent Constraint languages with the possibility to manage both local (and global) knowledge of agents. Being constraints defined as soft, the underlying system of preferences consists in an ordered monoid where it is possible to natively represent partially ordered and bipolar (i.e., positive/negative) preferences. Such a formalism further generalise the previous literature, while it also grants  a \emph{polynomial}  representation of soft constraints that can be exploited in  several applications. Variables, which support constraints, can either be private to a single agent, or in common with all of them, thus allowing  for different levels of visibility. We provide labelled and unlabelled reduction semantics and their correspondence, together with a labelled bisimulation equivalence.

@@

The fact that if we switch two steps and then switch them back, we may obtain a derivation that is not coincident with the original one has unpleasant consequences. For instance, if we restrict $\rho_1'$ and $\rho_3'$ to the first two steps, we get non-equivalent derivations $\delta_1;\delta_2 \not\shifteq \delta_1;\delta_2'$, but $\rho_1 \shifteq[\sigma] \rho_3$, with $\sigma$ mapping $\delta_1$ to $\delta_1$ and $\delta_2$ to $\delta_2'$.
%
Such a behaviour with respect to composition does not allow to obtain a causal semantics for graph transformation systems.
More precisely, it does  not allow to associate an event structure to a derivation.
%
Thus, the notion of sequential independence has to be refined. The point is that in the associated shift equivalence we want to forget who made the fusion, 
but we should not forget (as it happens in the example) that some fusions occurred.

@@

\bibliographystyle{splncs04}
%\bibliographystyle{plainurl}
\bibliography{Unstable}
